Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/overview.tex
stage-n => nth-stage

We can write a staged version of \texttt{qs}, called \texttt{qss},
whose code is shown in \ref{fig:qs-staged}, where the first-stage code
are typeset in red, and the second-stage code are typeset in blue.  To
obtain \texttt{qss} from \texttt{qs}, we first make the input list a
stage-1 and the rank a stage-2 argument.
%
More precisely the input list has type $\curr\mathrm{list}$ (an
integer list ``now''), the rank has type $\fut\rmint$ (an integer in
the ``future''), and the return has type $\fut\rmint$.
%
We can then obtain \texttt{qss} by annotating the body of \texttt{qs}
with the staging annotations \texttt{prev} and \texttt{next}, which
transition between a stage-1 and stage-2 code, and~$\texttt{g}$, which
marks ground terms.  
%
We also use a third function
\texttt{hold}, which can be implemented with \texttt{prev} and
\texttt{next}, for promoting a stage-1 integer to a stage-2 integer;
the signature of \texttt{hold} is
% 
\lstinline{1`hold : ^int -> $`2`int`}.

@@

The splitting algorithm scans the program code for first-stage
computations (which depend only on first-stage values) and separates
them into the function, \texttt{qs1} of the first pass. This function
performs the stage-1 computations and places the results into a
boundary data structure that both records the control flow and the
results from the first stage at relevant control-flow points.  More
specifically, the function \texttt{qs1} performs all the recursive
calls and evaluates all instances of the partition function, which
depend only on the input list.  The function produces a boundary data
structure that collects the results from all recursive calls along
with a tag that indicates the control branches taken.  Since
\texttt{qss} has a binary control structure (casing on the list), the
boundary is a binary tree.  To ensure that the trace contains the
necessary results to complete the execution in the second pass, where
second-stage values may be used, the splitting algorithm includes in
the nodes of the tree information such as the "pivot" used for
splitting and the size of the subtree at that note.  The resulting
tree is thus not just a tree but a binary search tree, keyed by the
pivot and augmented with "size" information.

>>>>

We can write a staged version of \texttt{qs}, called \texttt{qss},
whose code is shown in \ref{fig:qs-staged}, where the first-stage code
are typeset in red, and the second-stage code are typeset in blue.  To
obtain \texttt{qss} from \texttt{qs}, we first make the input list 
first-stage and the rank second-stage.
%
More precisely the input list has type $\curr\mathrm{list}$ (an
integer list ``now''), the rank has type $\fut\rmint$ (an integer in
the ``future''), and the return has type $\fut\rmint$.
%
We can then obtain \texttt{qss} by annotating the body of \texttt{qs}
with the staging annotations \texttt{prev} and \texttt{next}, which
transition between first-stage and second-stage code, and~$\texttt{g}$, 
which marks ground terms.  
%
We also use a third function
\texttt{hold}, which can be implemented with \texttt{prev} and
\texttt{next}, for promoting a first-stage integer to a second-stage integer;
the signature of \texttt{hold} is
% 
\lstinline{1`hold : ^int -> $`2`int`}.

@@

The splitting algorithm scans the program code for first-stage
computations (which depend only on first-stage values) and separates
them into the function, \texttt{qs1} of the first pass. This function
performs the first-stage computations and places the results into a
boundary data structure that both records the control flow and the
results from the first stage at relevant control-flow points.  More
specifically, the function \texttt{qs1} performs all the recursive
calls and evaluates all instances of the partition function, which
depend only on the input list.  The function produces a boundary data
structure that collects the results from all recursive calls along
with a tag that indicates the control branches taken.  Since
\texttt{qss} has a binary control structure (casing on the list), the
boundary is a binary tree.  To ensure that the trace contains the
necessary results to complete the execution in the second pass, where
second-stage values may be used, the splitting algorithm includes in
the nodes of the tree information such as the "pivot" used for
splitting and the size of the subtree at that note.  The resulting
tree is thus not just a tree but a binary search tree, keyed by the
pivot and augmented with "size" information.
